\pisodm: Building SOC Applications in the Presence of Non-Functional Requirements

### 2016-11-29 JournalPISODM2016/submitted/main.tex
journal version

\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}

\newcommand{\umberto}[1]{\marginpar{\textcolor{orange}{\textbf{Umberto: }#1}}}
\newcommand{\iumberto}[1]{\textcolor{orange}{\textbf{Umberto: }#1}}

@@

\LRH{K. Belhajjame et~al.}

@@

%
\authorA{Khalid Belhajjame}
%
\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
  E-mail: kbelhajj@googlemail.com}
%
%
\authorB{Valeria~de~Castro}
\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
  E-mail: valeria.decastro@urjc.es}
\authorC{Umberto~Souza~da~Costa}
\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
  E-mail: umberto@dimap.ufrn.br}

\authorD{Javier~A.~Espinosa-Oviedo}
\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
  E-mail: javiera.espinosa@gmail.com}

\authorE{Martin~A.~Musicante}
\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
  E-mail: mam@dimap.ufrn.br}

\authorF{Pl\'acido~A.~Souza~Neto}
\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
  E-mail: placido.neto@ifrn.edu.br}

\authorG{Genoveva~Vargas-Solar}
\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
  E-mail: genoveva.vargas@imag.fr}

\authorH{Jos\'e-Luis~Zechinelli-Martini}
\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
  E-mail: joseluis.zechinelli@udlap.mx}

%
%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
%
%\affA{Department of Biomedical Informatics, Emory University
%\newline
%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
%
%
%
%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
%\affB{Oracle \newline
 %New Jersey, USA}
%
%

@@

\begin{bio}
Khalid Belhajjame\dots Bio here.
\vs{9}

\noindent Valeria~de~Castro\dots Bio here.
\vs{9}

\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
Service-Oriented Computing and Cloud Computing.
\vs{9}

\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
\vs{9}

\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
his M.Sc and PhD in Computer Science at Universidade Federal de
Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
Brazil). He is part of the graduate program in Computer Science at
UFRN and holder of a research grant from CNPq (level 2). Martin was
associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
Martin A. Musicante's main current research topics are big data
processing languages, Web Services and cloud computing.
\vs{9}

\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
\vs{9}

\noindent Genoveva~Vargas-Solar\dots Bio here.
\vs{9}

\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
\end{bio}

@@

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010,SouzaNeto201684}. 
As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.

@@

\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue\martin{Placido: Mudar as cores... } meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules.
The conditions of each rule are evaluated.
In case of failure, the actions of the rule will be performed.
The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.

@@

\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03} or PEWS~\cite{Placido2010LTPD}.
%
PEWS~\cite{BHM06,Placido2010LTPD} is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:

@@

As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
transformed into a  {\sf Fork node}~\cite{valeriaThesis,placidoPhDThesis2012}.
If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.   
\martin{Nao entendo muito o algoritmo proposto aqui e o exemplo eh vago.}

%
A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
\martin{E as outras constraints?}

@@

{\sf Include} use case entities are transformed into an {\sf Action} sequence.
A {\sc Use case} element is transformed into an {\sf Action}. 
A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements. 
%
Details of these transformations can be found in~\cite{placidoPhDThesis2012}.

@@

\martin{Nesta secao e na seguinte tem transformacoes de constraints into apolicies e o inverso fica muito confuso.}

@@

To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.\martin{Is this a case study or a proof of concept?}
Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes, as well as to propose amendments to these practices.
The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.

@@

The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints of the \FlyingPig\ application.
In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
Company is the actor requesting a risk evaluation as a Broker is  responsible for coordinating  the evaluation process, assigning users to tasks as well as delegating tasks.
A User is an actor who answers questionnaires according to the current situation  of the Company.
The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
Each actor is associated to $\pi$-UseCases (white ovals in Fig.~\ref{fig:piUseCaseModel})
that describe the main functionality of the system.
The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-UseCases.
%%
Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).\martin{colored?}
Three types of constraints are defined: \textit{value}, \textit{business} or \textit{exceptional behavior}.
Constraints are identified by the word $<<$\textsf{constraint}$>>$ followed by its type.

>>>>

%\newcommand{\martin}[1]{\marginpar{\textcolor{magenta}{\textbf{Martin: }#1}}}
%\newcommand{\imartin}[1]{\textcolor{blue}{\textbf{Martin: }#1}}
%
%\newcommand{\umberto}[1]{\marginpar{\textcolor{orange}{\textbf{Umberto: }#1}}}
%\newcommand{\iumberto}[1]{\textcolor{orange}{\textbf{Umberto: }#1}}

@@

\LRH{  }%K. Belhajjame et~al.}

@@

%%
%\authorA{Khalid Belhajjame}
%%
%\affA{Universit\'e de Paris - Dauphine -- Paris, France\\
%  E-mail: kbelhajj@googlemail.com}
%%
%%
%\authorB{Valeria~de~Castro}
%\affB{Universidad Rey Juan Carlos -- M\'{o}stoles, Spain\\
%  E-mail: valeria.decastro@urjc.es}
%\authorC{Umberto~Souza~da~Costa}
%\affC{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
%  E-mail: umberto@dimap.ufrn.br}
%
%\authorD{Javier~A.~Espinosa-Oviedo}
%\affD{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
%  E-mail: javiera.espinosa@gmail.com}
%    
%\authorE{Martin~A.~Musicante}
%\affE{Universidade Federal do Rio Grande do Norte -- Natal, Brazil\\
%  E-mail: mam@dimap.ufrn.br}
%
%\authorF{Pl\'acido~A.~Souza~Neto}
%\affF{Instituto Federal de Educa\c{c}\~{a}o, Ci\^{e}ncia e Tecnologia do Rio Grande do Norte -- Natal, Brazil\\
%  E-mail: placido.neto@ifrn.edu.br}  
%
%\authorG{Genoveva~Vargas-Solar}
%\affG{CNRS, LIG-LAFMIA, Saint Martin d'H\`eres, France\\
%  E-mail: genoveva.vargas@imag.fr}  
%
%\authorH{Jos\'e-Luis~Zechinelli-Martini}
%\affH{Universidad de las Am\'ericas-Puebla, LAFMIA -- Cholula, Mexico\\
%  E-mail: joseluis.zechinelli@udlap.mx}
%
%
%%
%%\authorA{Fusheng Wang\footnote{Work done while working at Siemens Corporate Research.} }
%%
%%\affA{Department of Biomedical Informatics, Emory University
%%\newline
%%36 Eagle Row, Ste 589, Atlanta, GA 30322, USA}
%%
%%
%%
%%\authorB{\footnotesize Cristobal Vergara-Niedermayr\footnote{Work done while working at Siemens Corporate Research.}}
%%\affB{Oracle \newline
% %New Jersey, USA}
%%
%

@@

%\begin{bio}
%Khalid Belhajjame\dots Bio here.
%\vs{9}
%
%\noindent Valeria~de~Castro\dots Bio here.
%\vs{9}
%
%\noindent Umberto Costa received the BS degree in Computer Science from Federal University of Rio Grande do Norte in 1998. 
%He obtained the MS degree in Systems and Computation from Federal University of Rio Grande do Norte in 2000. 
%He received the PhD degree in Computer Science from University Federal de Minas Gerais in 2005. He is currently an associate professor in the Department of 
%Informatics and Applied Mathematics (DIMAp) at Federal University of Rio Grande do Norte. His academic interests include Programming Languages, 
%Service-Oriented Computing and Cloud Computing.
%\vs{9}
%
%\noindent Javier~A.~Espinosa-Oviedo\dots Bio here.
%\vs{9}
%
%\noindent Martin A. Musicante received his B.Sc at ESLAI, Argentina, in 1988 and
%his M.Sc and PhD in Computer Science at Universidade Federal de
%Pernambuco, Brazil, in 1990 and 1996, respectively. He is an associate
%professor at Universidade Federal do Rio Grande do Norte-UFRN (Natal,
%Brazil). He is part of the graduate program in Computer Science at
%UFRN and holder of a research grant from CNPq (level 2). Martin was
%associate researcher at LI - Universit\'e Fran\c{c}ois Rabelais Tours
%(2002-2011) and at LIFO - Universit\'e d'Orl\'eans, France, since 2008.
%Martin A. Musicante's main current research topics are big data
%processing languages, Web Services and cloud computing.
%\vs{9}
%
%\noindent Pl\'acido~A.~Souza~Neto\dots Bio here.
%\vs{9}
%
%\noindent Genoveva~Vargas-Solar\dots Bio here.
%\vs{9}
%
%\noindent Jos\'e-Luis~Zechinelli-Martini\dots Bio here.
%\end{bio}

@@

Beside the existence of these works, it is important to remark that authors continue emphasizing the limited attention of the model-driven proposals for service development to NFRs~\cite{Ameller201542,Ameller2010}.%,SouzaNeto201684}. 
As Ameller et al. realize in a recently published literature review, there are \textit{``Not many papers that presented a generic approach to deal with NFRs"}. Those authors also remark the lack of industrial use cases~\cite{Ameller201542}. 
Our method to develop service-oriented applications aims to cover this lack of proposals dealing with NFRs and gives support to real needs, as presented in a company related to the risk assessment industry.

@@

\item In order to represent constraint types associated to services compositions, we defined the me\-ta- classes {\sc Rule} and {\sc A-policy} (see blue
%\martin{Placido: Mudar as cores... } 
 meta-classes in the $\pi$-Serv\-ice\-Com\-po\-si\-tion meta-model in Figure \ref{fig:e-scomposition-metamodel}).
We model non-func\-tion\-al constraints by using the notion of {\em A-policy}~\cite{Espinosa-Oviedo2011a,CIC:eovszmc09c}.
An {\em A-policy} is defined by attributes and rules.
The conditions of each rule are evaluated.
In case of failure, the actions of the rule will be performed.
The meta-class {\sc Rule} represents event-condition-action rules where the {\sc Event} part represents the moment in which a constraint  is evaluated.
An {\em A-policy} defines variables and operations that can be shared by the rules and that can be used for expressing their Event and Condition parts.

@@

\textit{$\pi$-PEWS} provides con\-cepts for modelling service compositions.
Instances of this meta-model are textual descriptions of service compositions that can be translated into any service composition language, such as BPEL~\cite{bpel03}.% or PEWS~\cite{Placido2010LTPD}.
%
PEWS~\cite{BHM06}%,Placido2010LTPD}
 is a notation to express service compositions.
The language is based on the notion of Path Expressions~\cite{And79} and can easily be translated into any actual composition language, such as BPEL~\cite{bpel03}.
Figure~\ref{fig:PPEWSmetamodel} presents the $\pi$-{\sc Pews} meta-model, where we identify classes to describe:

@@

As defined by the $\pi$-ServiceProcess  meta-model (Figure~\ref{fig:CIM:serviceprocessmetamodel}) an {\sc Activity Service} consists of a composition of  entities of type {\sc Action}. 
Thus, every {\sf $\pi$-use case} is transformed into an {\sf Action} of the target $\pi$-Service Process model.  
Every {\sf Extend} relationship identified in a $\pi$-Use Case model is
transformed into a  {\sf Fork node}~\cite{valeriaThesis}.%,placidoPhDThesis2012}.
If the {\sf Extend} relationship concerns just one {\sf use case}, it is transformed into an {\sf Action} inside one flow of the fork node. 
Otherwise, several  {\sf use cases} are transformed into different {\sf Actions} that belong to different flows departing from the fork node.   
%\martin{Nao entendo muito o algoritmo proposto aqui e o exemplo eh vago.}

%
A {\sf Constraint} associated to a {\sf Use Case}  is transformed into an  {\sf Assertion}.  
The set of resulting assertions  are grouped into a {\sf Contract}.
Constraints are transformed according to their type:
{\sc Business} constraints and {\sc Value} constraints with the {\sf  isExceptionalBehaviour} at\-tri\-bute set to false are transformed into {\sf Assertion}s;
{\sc Value } constraints with the {\sf  isExceptionalBehaviour} attribute set to true are transformed into {\sf Exceptional behavior}s.
%\martin{E as outras constraints?}

@@

{\sf Include} use case entities are transformed into an {\sf Action} sequence.
A {\sc Use case} element is transformed into an {\sf Action}. 
A set of $n$ {\sc Use cases} is transformed into an  $n-1$ {\sf Object flow} elements. 
%
%Details of these transformations can be found in~\cite{placidoPhDThesis2012}.

@@

%\martin{Nesta secao e na seguinte tem transformacoes de constraints into apolicies e o inverso fica muito confuso.}

@@

To validate the applicability of our method we have developed a case study concerning risk assessment for financial companies as implemented by the ORCA System\footnote{The ORCA System is a trademark of GCP Global (www.gcpglobal.com).}.%\martin{Is this a case study or a proof of concept?}
Risk assessment is implemented by an interactive business process based on the exchange of questionnaires used to evaluate the risks implied by the client business practice.
Examples of business practices are: the conditions and protocols used to perform confidential transactions, the physical security to access reserved areas (such as computing server installations).
The information gathered by the questionnaires is used to evaluate whether there are risky practices within the business processes, as well as to propose amendments to these practices.
The ultimate goal of the risk assessment is to determine a degree of compliance to existing standards.
By analyzing the questionnaires, ORCA detects risky practices, proposes solutions and triggers further assessment processes to ensure that the solutions have been implemented.

@@

The $\pi$-UseCase model shown in Figure~\ref{fig:piUseCaseModel} describes the features and constraints of the \FlyingPig\ application.
In this model, three actors are identified: \textit{Company}, \textit{User} and \textit {Broker} represented as stick figures.
Company is the actor requesting a risk evaluation as a Broker is  responsible for coordinating  the evaluation process, assigning users to tasks as well as delegating tasks.
A User is an actor who answers questionnaires according to the current situation  of the Company.
The User also produces evidence to support facts and performs the necessary amendments to improve the results of the risk assessment.
Each actor is associated to $\pi$-UseCases (white ovals in Fig.~\ref{fig:piUseCaseModel})
that describe the main functionality of the system.
The $\pi$-UseCase model for \FlyingPig\ defines six $\pi$-UseCases.
%%
Each $\pi$-UseCase may be associated to (non-functional) constraints (coloured ovals in Fig.~\ref{fig:piUseCaseModel}).%\martin{colored?}
Three types of constraints are defined: \textit{value}, \textit{business} or \textit{exceptional behavior}.
Constraints are identified by the word $<<$\textsf{constraint}$>>$ followed by its type.
